How geographic access to care shapes disease burden: the current impact of post-exposure prophylaxis and potential for expanded access to prevent human rabies deaths in Madagascar

### 2021-01-13 analysis/paper/minor_revision_1/response_reviewers.Rmd

We thank the editor and the reviewers for their constructive feedback and close reading of the manuscript and have revised our manuscript accordingly. Both a document with track changes (of both the manuscript & the supplement) and clean separate documents have been submitted, and we have also verified our figures meet the requirements per the PACE tool. The reviewer comments are below with our response in italics.

### Reviewer 1

-   The authors performed the well-designed epidemiological analysis; given the distance from the residence to the PEP accessibility and assess the rabies risk. There is only once concern arising how much the prevalence (or exposure to residence) of rabid dogs in each area was included into this study. The authors also pointed out the uncertainty of underlying rabies expose according to the sensitivity analysis. This should be more clearly describe in the manuscript, with a validity of the corresponding parameters used in the current study.

    *We thank the reviewer for their helpful feedback. We have revised the methods and results section based on this and another reviewer's feedback. The methods section describing the decision tree now forefronts the parameter estimates with their sources. The results section now maps onto the major findings of each of our analyses, and hopefully clarifies the results of each section. We have also added a sentence to the discussion to regarding the validity of our assumptions (see lines 624 - 626).* <!-- Done -->

@@

-   Overall, the statements in Conclusion part is adequate and convince readers.

    *Thank you.* <!-- Done -->

-   In Lines 523-525, the authors discuss the potential risk factor of human rabies in Madagascar. These are very general and commonly argued in anywhere. Don't the authors can find the supportive reasons or data for these assumptions? Specifically, box1 may provide the example of association between area and awareness to the risk. We would like to know how these gaps may arise.

    *We have reorganized this discussion point, and and point to a recent KAP study in the Moramanga District that identified such barriers to PEP (see lines 644 - 649). We also discuss the challenges in identifying these risk factors and link this to the case studies in Box 1 (lines 652 - 656).* <!-- Done -->

@@

### Reviewer 2

-   The authors describe an approach to model the efficacy of different PEP distribution scenarios scenarios interesting and new. The approach is based on several steps and each steps is complex and relies on a set of assumptions. Because it would be difficult to collect empirical data (e.g. in a cluster randomized trial), I think the study is worth publishing. I appreciate that a lot of details and supplementary information is provided and that the source code and data is public. My main critic points are: i) due to the complexity and amount and diversity of information (the appendix alone has 47 pages) the authors should guide the reader better through the manuscript. Some suggestions are provided below, ii) the results are based on many assumptions -- I am not sure if the authors are not a bit too optimistic with their approach. I think the next step should be to try to validate the model before "Our framework could be used to guide PEP expansion and improve targeting of interventions."

    *We thank the reviewer for their helpful feedback and have revised the manuscript accordingly (see specific points regarding revisions to the methods and results in response to points below). Additionally, we have revised the text in the abstract (lines 63 - 68) and in the conclusion (line 764) to state that in the context of countries becoming eligible for GAVI expansion, this framework may be a useful way to move forward in the absence of baseline data on human rabies risk, and surveillance should be improved to evaluate the impacts of expanding PEP. We also added a point that in other settings, addressing other barriers to PEP access may be more effective (i.e. direct costs of PEP in settings that charge for PEP) (see lines 701 - 704 in the 'Broader context' section of the discussion).* <!-- Done -->

@@

-   Ideas to guide the reader a bit better through the manuscript: - One weakness of the paper is that the reader has to jump back and forth to get all the information. Different parameters are introduced over 5 pages. Maybe the method section could be rearranged and a bit sharpened. E.g. p[rabid] is introduced in line 201. In line 208 (a figure legend) it is explained that R[rabid] is constrained. In line 225 it is explained which range was used to express uncertainty about P[rabies], in line 229 the constraining formula is provided and finally in Table 1 the values distribution and reference for this parameter is provided. I don't know if the constraint is so important that it has to be mentioned to this extend. ;Maybe a footnote in table1 would be sufficient. If not, there is certainly a better way to arrange the method section that it can be introduced at only one (maybe 2) occasion.

    *We have revised the methods section accordingly (lines 242 - 277). Now all parameters are introduced in Table 1 with their sources and details, and Figure 1 describes how they are used in the decision tree framework. We have moved Table 1 up to the top of the section and describe the constraint on p[rabid] in a footnote and removed this from the figure description. The minor details about both p[rabid] and rho[max] are now described in the `Description` column of the table.* <!-- Done -->

@@

-   Same with bite incidence. The result section shows some graphs and a lot of information related to model fit. But the reader is likely more interested in the estimated mean incidence. But this value is reported is not in this but in the next paragraph.

    *For the results section, we have revised each section to better state the main finding up front and focus on details in subsequent paragraphs. We combined the sections describing the relationship between bite incidence and travel times and moved some of the details regarding bite incidence estimation to the Supplementary Appendix section S2 and some of the details on model fit and comparisons to section S3.* <!-- Done -->

-   Terminology could be simplified. E.g. using consistently either commune or CSB2 in results and discussion to describe this level.

    *We have added a short glossary to guide the reader through the manuscript (lines 42 - 60) and also changed all mentions of CSB-II to primary clinic and CSB-I to secondary clinic, and use the term clinics throughout to refer to health facilities.* <!-- Done -->

@@

-   The most important result is "Estimating the impact of PEP provisioning." I think it would be good to have these results summarized in a table. This table could include/replace table 2.

    *We include an estimate of the deaths averted under current PEP provisioning in Table 2 and clarified the column names to make this more clear.* <!-- Done -->

-   Out-of-stock is briefly mentioned in this section but I think it is worth to discuss also that the supply chain is getting much more complicated if PEP hast o be delivered to much more clinics.

    *We expand on this later in the discussion, and we have rephrased the sentence in the discussion to point directly to complications in the supply chain. We also point to potential solutions for managing these supply chains (i.e. piggybacking on EPI programs, see lines 714 - 730).* <!-- Done -->

@@

-   Abstract - After reading the abstract I had no clear impression what the authors exatly did, e.g. ehat kind of data they used to estimate travel time, bite incidence and rabies deaths.

    *We have revised the abstract to more clearly describe the data and methods we used (lines 42 - 46).* <!-- Done -->

@@

-   "expanding PEP to one clinic per district could reduce deaths by 19%" Please add the number of districts (114) or the number of additional clinics (83).

    *Done.* <!-- Done -->

-   Authors summary " but our results suggest that expansion alone will not eliminate deaths." Well, how could PEP eliminate human rabies deaths? It does not interrupt transmission. I think this statement should be rephrased.

    *We have rephrased this to point to the fact that PEP expansion alone will be unlikely to result in the goal of zero human rabies deaths by 2030, and are trying to communicate to policy makers that countries will need to invest in dog vaccination if this goal is to be achieved (lines 84 - 85).* <!-- Done -->

@@

-   Travel time: The authors used geographical information to estimate theoretical travel time as an proxi for access. In addition they tried to validate the approach by comparing the estimates with travel time reported by patients and by travel time reported by the Institute Pasteur. However, very basic information is lacking, e.g. I could even not find the number of patients interviewed and it remains unclear what " driving times collected by IPM during field missions " actually means. Of course, reported and estimated travel times are correlated -- both are associated with distance -- but if I look at figure 2c or the appendix the variation is remarkable. Some estimate of fit would be nice. Maybe mean/median residual or something similar. By the way: I don't think the shaded areas (CIs around the regression lines) does not add a lot tot he interpretation. I would remove them because the points below are difficult to see.

    *We have clarified both the methods and added sample sizes(lines 162 - 180), as well as the results the results (lines 356 - 382) based on the reviewer's helpful feedback. We point to some reasons for the observed variance in the results (lines 360 - 363) and in the discussion (lines 603 - 617), and a co-author and others are working on developing improved friction surfaces that could help better capture travel times based on Madagascar specific data. We also touch briefly on the issues with the data we used to 'ground-truth' travel times from the friction surface in the discussion (lines 614 - 615). See Table S1.1 for R^2^ estimates for the linear models we used to compare the metrics, which we now also report in the main results text (line 381). We removed the CIs around the regression lines for both Fig 2C and Fig S1.4.* <!-- Done -->

@@

    *Thanks for **pointing** this out, fig S2.2 and it's caption have been revised accordingly, with S2.2A showing the number of forms submitted for each clinic with excluded periods shown in grey and S2.2B showing the four point estimates of reporting and the line showing the range.* <!-- Done -->

    #### Minor points :

-   Is the overdisperion parameter presented in Fig 4 the estimate or log(estimate)? If it is the estimate, an estimate below 1 is rather unusual.

    *The overdispersion estimate is not on a log scale, but we incorrectly describe it's form in the methods and have corrected it. The overdispersion parameter is the standard deviation of a random variable with mean 0 (see corrected lines 234 - 235) and it is accordingly scaled by the offset (population size). This means that the estimated overdispersion is actually quite large, as described in the text and vizualised in the figures. For example, given a travel time of zero with a population of 100,000 and taking approximately the mean of the posteriors, we would predict exp(-5.5 +/- 0.9)100000 = 170 - 1000 bites per 100,000 persons. We apologize for the error in the text and thank the reviewer for catching it!* <!-- Done -->

@@

-   Delete in line 132 'the'

    *Corrected* <!-- Done -->

-   Fig. 1 with its description is confusing. I would rather recommend a list with the variables and their description.

    *We have revised this section based on this and another reviewers feedback (lines 242 - 277). Briefly, we moved the table with the parameters and description to the top of the section, use the figure to show how these parameters were used, and additional choices regarding parameter ranges in the subsequent paragraph.* <!-- Done -->

@@

-   Line 234, can this really be assumed? A baseline study in West Africa demonstrated, that only half of the bite victims complete PEP. However, based on the information regarding the high patient compliance in Madagaskar (line 539), I think you can make this assumption here.

    *Thank you for pointing this out, we have moved the reference from line 539 to the methods (now line 275) to ground this assumption up front.* <!-- Done -->

@@

-   Remove a % in line 493

    *Corrected.* <!-- Done -->

-   This work nicely demonstrates that even with provisioning PEP all over the country, there will still be rabies deaths in the country. So besides the provision of PEP for humans, the disease needs to be eliminated in the animal reservoir.

    *Thank you, we've also revised these points based on the other reviewers feedback to address that PEP alone is unlikely to get countries to the goal of 'Zero by 30' (the WHO goal of zero human rabies deaths from dog-mediated rabies by 2030; see lines 84 - 85).* <!-- Done -->

### 2021-01-13 analysis/paper/manuscript.Rmd
updating with figs & files and making paper instructions

From GPS locations of the 31 ARMCs we estimated the travel time to the nearest clinic at an approximately 1 x 1 km scale as described in [@weissGlobalMapTravel2018]. We then extracted the mean and population-weighted mean travel times for the district and commune (the administrative unit below the district), and Euclidean distance, i.e. the minimum distance from the administrative unit centroid to any ARMC. We used shapefiles from the UN Office for the Coordination of Humanitarian Affairs for the administrative boundaries (as of October 31, 2018). To see which metric best predicted ground-truthed travel time data, we compared travel times and distance estimates to driving times collected by IPM during field missions, i.e. time it took to travel by car between two locations excluding break times (N = `r `nrow(ttime_driving)`), and patient reported travel times from a subset of Moramanga ARMC patients (N = `r patient_nobs`), see Fig S1.2 for raw data) by seeing which worked best in terms of a linear model.

@@

## Modeling human rabies deaths {#methods-burden}
We estimate rabies deaths as a function of the number of bites predicted by our model and estimates of endemic rabies exposure incidence using an adapted decision tree framework. Table 1 lists all parameter values and their sources. Fig 1 describes how these parameters are used in the decision tree and the key outputs ($A_i$, deaths averted by PEP, and $D_i$, deaths due to rabies).

@@

@@

| **Parameter**   | **Value**        | **Description**	                 | **Source**|
|:----------------|:----------------:|:----------------------------------|:---------:|
| $B_i$ |	Function of travel time to closest ARMC | Modeled estimates of reported bite incidence	| Bayesian regression model (see Methods) |
| $E_i$ |	Triangular(a = 15, b = 76, c = 42)| Annual exposures per 100,000 persons	| [@whorabiesmodelingconsortiumPotentialImpactImproved2019; @rajeevHealthcareUtilizationProvisioning2019], see Fig S4.1|
| $p_{rabid}$	| Triangular(a = 0.2, b = 0.6, c = 0.4) |	Proportion of reported bites that are rabies exposures^1^ | [@rajeevHealthcareUtilizationProvisioning2019] |
| $\rho_{max}$ | `r round(rho_max, 2)`	| The maximum reporting possible for any location; data from the commune closest to the Moramanga ARMC (average of 3.12 minutes travel time to the clinic)| [@rajeevHealthcareUtilizationProvisioning2019] |
| $p_{death}$ |	0.16 | The probability of death given a rabies exposure	| [@changaluchaNeedImproveAccess2018]|

^1^ $p_{rabid}$ is constrained so that rabid reported bites cannot exceed the total expected number of rabies exposures ($E_i$) or maximum reporting in a given simulation ($\rho_max$).

@@

We simulate expansion first to each district (N = `r nrow(mada_districts)`) and then to each commune in the country for all communes with a clinic. We select the primary clinic (primary health facility, see Glossary) in the highest density grid cell of the administrative unit as candidates for expansion. For the `r comms_no_csb2` communes without a primary clinic, we chose the secondary clinic (secondary health facility, see Glossary) in the highest density grid cell. `r comms_no_clins` communes lacked any health facilities. Finally, we explore a scenario where all additional primary clinics (totalling `r nrow(csbs_max)`) provision PEP.

@@

Bite incidence estimates generally increased with decreasing weighted travel times at both scales, although there was considerable variation between catchments for the magnitude of this relationship (Fig 3C and D). After additionally excluding any year with less than 25% of forms submitted, our final dataset consisted of estimates of average bite incidence for `r nrow(district_bites)` of `r nrow(mada_districts)` districts (Fig 3C), and `r nrow(mora_bites)` communes within the catchment of the Moramanga District (Fig 3D). For the national data, there were two outliers, Toamasina II (the sub-urban district surrounding the city of Toamasina) and Soanierana Ivongo, with higher bite incidence when compared to other districts with similar travel times. While the estimates from the Moramanga data showed higher reported incidence at low travel times at the commune level compared to the district estimates, when aggregated to the district, bite incidence estimates fell within the ranges observed from the national dataset.

@@

## PEP currently averts the majority of estimated human rabies deaths, but burden remains high in areas with poor access to PEP. {#results-burden}
In general, the incidence of rabies deaths increases with travel times to clinics, and there is significant sub-national variation when deaths are modeled at the district and commune scale, with the least accessible communities having most deaths (Fig 5B & C). Under the current system of 31 ARMCs in Madagascar, we estimate that use of PEP prevents approximately `r round(natl_base$averted_mean[natl_base$scale == "Commune"], -2)` (95% PI: `r round(natl_base$averted_lower[natl_base$scale == "Commune"], -2)` - `r round(natl_base$averted_upper[natl_base$scale == "Commune"], -2)`) deaths due to rabies each year. Overall, we estimate close to `r round(natl_base$deaths_mean[natl_base$scale == "Commune"], -2)` rabies deaths (95% PI: `r round(natl_base$deaths_lower[natl_base$scale == "Commune"], -2)` - `r round(natl_base$deaths_upper[natl_base$scale == "Commune"], -2)`) annually in Madagascar. Our estimates vary only slightly depending on the scale of the model (Table 2), but disaggregating deaths to the commune level shows considerable variation in predicted burden within districts (Fig 5A).

@@

### *Fig 6. Impact of expanded PEP access on deaths, deaths averted and vial demand.*
*(A) Decrease in deaths due to rabies, (B) increase in total # of vials as additional ARMC are added at the national level , and (C) increase in vials needed per death averted based on the two models of reported bites. Lines are the mean of 1000 simulations with envelopes representing 95% prediction intervals. The points show when all additional primary clinics and secondary clinics (N = `r max_added`) clinics have been added).*

@@

Our burden estimates were most sensitive to assumptions about rabies exposure incidence, drawn from studies in the Moramanga District [@rajeevHealthcareUtilizationProvisioning2019] and elsewhere [@whorabiesmodelingconsortiumPotentialImpactImproved2019]. As incidence of rabies exposures varies over time and space [@hampsonSynchronousCyclesDomestic2007; @hampsonTransmissionDynamicsProspects2009], we incorporated uncertainty into our estimates, but we did not find qualitative differences in the effects of travel times on rabies deaths. While we rely on a number of assumptions, they are based on data specific to Madagascar or from other similar settings and consistent with estimates from the literature more broadly (see Table 1 and section S4). Our simplifying assumptions regarding patient compliance, which is generally high in Madagascar [@rajeevHealthcareUtilizationProvisioning2019], and on complete efficacy of PEP are unlikely to greatly influence our burden estimates [@changaluchaNeedImproveAccess2018]. Likewise, we do not account for differential risk for severely exposed patients not receiving Immunoglobulins (RIG), which is only available at IPM in Antananarivo, but recent studies show that even in the absence of RIG, PEP is extremely effective [@whorabiesmodelingconsortiumPotentialImpactImproved2019]. We also assume that all ARMC reliably provision PEP, but a 2019 KAP survey reported clinics experiencing stock-outs [@leblancclaireRabiesMadagascarThreepronged2019].

We assumed geographic access to PEP was the primary driver of health-seeking behavior, but socioeconomic status, education and awareness about rabies [@etheartEffectCounsellingHealthcareseeking2017; @hampsonRabiesExposuresPostExposure2008; @barbosacostaInfluencePovertyRabies2018; @samboKnowledgeAttitudesPractices2014] all play a role.  Most ARMC also charge fees (from 0.50 - 3.00 USD for consultations, wound treatment, etc. [@leblancclaireRabiesMadagascarThreepronged2019]) which may also act as barriers to PEP access. In Madagascar, where PEP is free-of-charge, the main cost to patients is transport and time lost. More remote communities tend to be of lower socioeconomic and educational status [@peters2008poverty], so travel time may be a proxy for these correlated variables. 
Significant overdispersion in the data that cannot be explained by travel times suggests that clinic-level variation (e.g. vaccine availability and charges) and regional differences (e.g. dog populations, outbreaks, awareness) further influence health-seeking behavior and vaccine demand. Although our estimates could be improved with better data on rabies incidence, health-seeking behavior, and PEP provisioning, predicting PEP impacts will remain challenging given the complex interactions between socioeconomic factors, access to and quality of care, and human behavior, as illustrated by the case studies in Box 1.

@@

However, our results suggest that PEP expansion alone cannot prevent the majority of rabies deaths, and even given maximal access, achieving ‘the last mile’, preventing deaths in the most remote populations, will require disproportionate resources [@delriovilasTribulationsLastMile2017]. To achieve ‘Zero by 30’, mass dog vaccination will be key to interrupting transmission, and eliminating deaths. Integrated Bite Case Management (IBCM) uses bite patient risk assessments to determine rabies exposure status, guide PEP administration, and trigger investigations of rabid animals, potentially identifying other exposed persons [@lechenneImportanceParticipatoryIntegrated2017; @rajeevHealthcareUtilizationProvisioning2019; @undurragaCostEffectivenessEvaluationNovel2017]. IBCM is one way to manage PEP effectively [@rysavaPathRabiesElimination2018] and as it relies on exposed persons reporting to clinics, expanding PEP access could strengthen this surveillance framework. These same issues of access, however, apply to both dog vaccination and surveillance, and understanding spatial heterogeneities will be critical to determining how control and prevention interventions can be best implemented [@fergusonHeterogeneitySpreadControl2015; @metcalfImplicationsSpatiallyHeterogeneous2013].

>>>>

From GPS locations of the 31 ARMCs we estimated the travel time to the nearest clinic at an approximately 1 x 1 km scale as described in [@weissGlobalMapTravel2018]. We then extracted the mean and population-weighted mean travel times for the district and commune (the administrative unit below the district), and Euclidean distance, i.e. the minimum distance from the administrative unit centroid to any ARMC. We used shapefiles from the UN Office for the Coordination of Humanitarian Affairs for the administrative boundaries (as of October 31, 2018). To see which metric best predicted ground-truthed travel time data, we compared travel times and distance estimates to driving times collected by IPM during field missions, i.e. time it took to travel by car between two locations excluding break times (N = `r `nrow(ttime_driving)`), and patient reported travel times from a subset of Moramanga ARMC patients (N = `r patient_nobs`), see Fig S1.2 for raw data) by seeing which worked best to predict estimated travel times in a linear model.

@@

## Modeling human rabies deaths {#methods-burden}

@@

We estimate rabies deaths as a function of the number of bites predicted by our model and estimates of endemic rabies exposure incidence using an adapted decision tree framework. Table 1 lists all parameter values and their sources. Fig 1 describes how these parameters are used in the decision tree and the key outputs ($A_i$, deaths averted by PEP, and $D_i$, deaths due to rabies).

@@

| **Parameter**   | **Value**        | **Description**	                 | **Source**|
|:----------------|:----------------:|:----------------------------------|:---------:|
| $B_i$ |	Function of travel time to closest ARMC | Modeled estimates of reported bite incidence	| Bayesian regression model (see Methods) |
| $E_i$ |	Triangular(a = 15, b = 76, c = 42)| Annual exposures per 100,000 persons	| [@whorabiesmodelingconsortiumPotentialImpactImproved2019; @rajeevHealthcareUtilizationProvisioning2019], see Fig S4.1|
| $p_{rabid}$	| Triangular(a = 0.2, b = 0.6, c = 0.4) |	Proportion of reported bites that are rabies exposures^1^ | [@rajeevHealthcareUtilizationProvisioning2019] |
| $\rho_{max}$ | `r round(rho_max, 2)`	| The maximum reporting possible for any location; data from the commune closest to the Moramanga ARMC (average of 3.12 minutes travel time to the clinic)| [@rajeevHealthcareUtilizationProvisioning2019] |
| $p_{death}$ |	0.16 | The probability of death given a rabies exposure	| [@changaluchaNeedImproveAccess2018]| 
- ^1^ $p_{rabid}$ is constrained so that rabid reported bites cannot exceed the total expected number of rabies exposures ($E_i$) or maximum reporting in a given simulation ($\rho_{max}$).

@@

We simulate expansion first to each district (N = `r nrow(mada_districts)`) and then to each commune in the country for all communes with a clinic. We select the primary clinic (primary health facility, see Glossary) in the highest density grid cell of the administrative unit as candidates for expansion. For the `r comms_no_csb2` communes without a primary clinic, we chose the secondary clinic (secondary health facility, see Glossary) in the highest density grid cell. `r comms_no_clins` communes lacked any health facilities. Finally, we explore a scenario where all additional primary clinics (totaling `r nrow(csbs_max)`) provision PEP.

@@

Bite incidence estimates generally increased with decreasing weighted travel times at both scales, although there was considerable variation between catchments for the magnitude of this relationship (Fig 3C and D). After additionally excluding any year with less than 25% of forms submitted, our final dataset consisted of estimates of average bite incidence for `r nrow(district_bites)` of `r nrow(mada_districts)` districts (Fig 3C), and `r nrow(mora_bites)` communes within the catchment of the Moramanga District (Fig 3D, see Supplementary Appendix section S2 for more details). For the national data, there were two outliers, Toamasina II (the sub-urban district surrounding the city of Toamasina) and Soanierana Ivongo, with higher bite incidence when compared to other districts with similar travel times. While the estimates from the Moramanga data showed higher reported incidence at low travel times at the commune level compared to the district estimates, when aggregated to the district, bite incidence estimates fell within the ranges observed from the national dataset.

@@

## PEP currently averts the majority of estimated human rabies deaths, but burden remains high in areas with poor access. {#results-burden}
In general, the incidence of rabies deaths increases with travel times to clinics, and there is significant sub-national variation when deaths are modeled at the district and commune scale, with the least accessible communities having most deaths (Fig 5B & C). Under the current system of 31 ARMCs in Madagascar, we estimate that use of PEP prevents approximately `r round(natl_base$averted_mean[natl_base$scale == "Commune"], -2)` (95% PI: `r round(natl_base$averted_lower[natl_base$scale == "Commune"], -2)` - `r round(natl_base$averted_upper[natl_base$scale == "Commune"], -2)`) deaths due to rabies each year. Overall, we estimate close to `r round(natl_base$deaths_mean[natl_base$scale == "Commune"], -2)` rabies deaths (95% PI: `r round(natl_base$deaths_lower[natl_base$scale == "Commune"], -2)` - `r round(natl_base$deaths_upper[natl_base$scale == "Commune"], -2)`) annually in Madagascar. Our estimates vary only slightly depending on the scale of the model (Table 2), but disaggregating deaths to the commune level shows considerable variation in predicted burden within districts (Fig 5A).

@@

### *Fig 6. Impact of expanded PEP access on deaths, deaths averted and vial demand.*
*(A) Decrease in deaths due to rabies, (B) increase in total # of vials as additional ARMC are added at the national level, and (C) increase in vials needed per death averted based on the two models of reported bites. Lines are the mean of 1000 simulations with envelopes representing 95% prediction intervals. The points show when all additional primary clinics and secondary clinics (N = `r max_added`) clinics have been added).*

@@

While we rely on a number of assumptions, they are based on data specific to Madagascar or from other similar settings and consistent with estimates from the literature more broadly (see Table 1 and section S4). Our burden estimates were most sensitive to assumptions about rabies exposure incidence, drawn from studies in the Moramanga District [@rajeevHealthcareUtilizationProvisioning2019] and elsewhere [@whorabiesmodelingconsortiumPotentialImpactImproved2019]. As incidence of rabies exposures varies over time and space [@hampsonSynchronousCyclesDomestic2007; @hampsonTransmissionDynamicsProspects2009], we incorporated uncertainty into our estimates, but we did not find qualitative differences in the effects of travel times on rabies deaths. Our simplifying assumptions regarding patient compliance, which is generally high in Madagascar [@rajeevHealthcareUtilizationProvisioning2019], and on complete efficacy of PEP are unlikely to greatly influence our burden estimates [@changaluchaNeedImproveAccess2018]. Likewise, we do not account for differential risk for severely exposed patients not receiving Immunoglobulins (RIG), which is only available at IPM in Antananarivo, but recent studies show that even in the absence of RIG, PEP is extremely effective [@whorabiesmodelingconsortiumPotentialImpactImproved2019]. We also assume that all ARMC reliably provision PEP, but a 2019 KAP survey reported clinics experiencing stock-outs [@leblancclaireRabiesMadagascarThreepronged2019].

We assumed geographic access to PEP was the primary driver of health-seeking behavior, but socioeconomic status, education and awareness about rabies [@etheartEffectCounsellingHealthcareseeking2017; @hampsonRabiesExposuresPostExposure2008; @barbosacostaInfluencePovertyRabies2018; @samboKnowledgeAttitudesPractices2014] all play a role. For example, most ARMC also charge fees (from 0.50 - 3.00 USD for consultations, wound treatment, etc. [@leblancclaireRabiesMadagascarThreepronged2019]) which may also act as barriers to PEP access. In Madagascar, where PEP is free-of-charge, the main cost to patients is transport and time lost. More remote communities tend to be of lower socioeconomic and educational status [@peters2008poverty], so travel time may be a proxy for these correlated variables. 
Significant overdispersion in the data that cannot be explained by travel times suggests that clinic-level variation (e.g. vaccine availability and charges) and regional differences (e.g. dog populations, outbreaks, awareness) further influence health-seeking behavior and vaccine demand. Although our estimates could be improved with better data on rabies incidence, health-seeking behavior, and PEP provisioning, predicting PEP impacts will remain challenging given the complex interactions between socioeconomic factors, access to and quality of care, and human behavior, as illustrated by the case studies in Box 1.

@@

However, our results suggest that PEP expansion alone cannot prevent the majority of rabies deaths, and even given maximal access, achieving ‘the last mile’, preventing deaths in the most remote populations, will require disproportionate resources [@delriovilasTribulationsLastMile2017]. To achieve ‘Zero by 30’, mass dog vaccination will be key to interrupting transmission and eliminating deaths. Integrated Bite Case Management (IBCM) uses bite patient risk assessments to determine rabies exposure status, guide PEP administration, and trigger investigations of rabid animals, potentially identifying other exposed persons [@lechenneImportanceParticipatoryIntegrated2017; @rajeevHealthcareUtilizationProvisioning2019; @undurragaCostEffectivenessEvaluationNovel2017]. IBCM is one way to manage PEP effectively [@rysavaPathRabiesElimination2018] and as it relies on exposed persons reporting to clinics, expanding PEP access could strengthen this surveillance framework. These same issues of access, however, apply to both dog vaccination and surveillance, and understanding spatial heterogeneities will be critical to determining how control and prevention interventions can be best implemented [@fergusonHeterogeneitySpreadControl2015; @metcalfImplicationsSpatiallyHeterogeneous2013].
